Context-dependent consequences of lagged effects in demographic models

### 2024-11-07 docs/ipm_comparison_ms.qmd
update to ms

---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 11pt
    documentclass: article
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        \usepackage[sc]{mathpazo} %Like Palatino with extensive math support
        \usepackage[utf8]{inputenc}

@@

@@

1.  Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.
2.  Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
3.  While recent mathematical and statistical developments have expanded the scope of these applications [e.g., @brooksStatisticalModelingPatterns2019; @reesEvolvingIntegralProjection2016], a number of technical challenges remain.
4.  While many of these relate to general issues surrounding model development and validation [e.g., @williams2012avoiding], others stem from the fundamental challenge of using mathematical models - albeit highly advanced ones - to describe and analyze complex biological systems.

@@

1.  Once such biological challenge is the implementation of alagged effects. It has long been recognized that there is the potential for lagged effects.
2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
3.  The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].

@@

1.  Lagged effects could be different in different habitat types.
2.  This could be a big reason for differences between habitats.
3.  Could also be important for stochastic vs. deterministic models.
4.  Here we...
5.  We conducted these analyses with both deterministic and stochastic IPMs.

@@

\`\`\`\`

>>>>

---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    documentclass: article
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        \usepackage[sc]{mathpazo} %Like Palatino with extensive math support
        \usepackage[utf8]{inputenc}
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \setlength{\parindent}{0pt}

@@

\bigskip

**Keywords:** demography, environmental stochasticity, integral projection models, lagged effects, structured population models, population dynamics

\bigskip

**Manuscript elements:** Figure\~1, figure\~2, table\~1, appendix\~A..

\bigskip

**Manuscript type:** e-note

@@

1.  Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.
2.  Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
3.  Mathematical and statistical advances have rapidly expand the scope of questions that can be addressed with these models [e.g., @brooksStatisticalModelingPatterns2019; @reesEvolvingIntegralProjection2016].
4.  However, there are still several fundamental questions for which these models are not yet readily applied, or for which it is unclear how potential alternative approaches will shape model results [e.g., @williams2012avoiding].

<!-- Not quite right yet...maybe:  Others, however, reflect the difficulty of describing complex biological processes with the mathematical and statistical frameworks used for demographic models;  models may be extremely difficult to parameterize, limited in  -->

<!-- recent advances greatly expanded the scope of questions that can be addressed. -->

@@

1.  One example of such a process is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects*.
2.  Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions.
3.  For instance, environmental conditions during juvenile development can shape the expression of traits (e.g., defensive spikes on *Daphnia*) that determine adult survival.
4.  Alternatively, the physiological mechanisms responsible for a vital rate can take an extended period of time to complete (e.g., bud formation in \_\_\_\_\_ can begin 1-2 years before flowers appear).
5.  Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).
6.  Although these lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their impacts remain poorly understood for two primary reasons.
7.  The first is the empirical challenge of designing studies to quantify lagged responses and evaluate their potential drivers [@kussEvolutionaryDemographyLonglived2008a].
8.  The second is the technical challenge of incorporating these complex or poorly understood biological processes in demographic models; doing often renders them less parameterizable, tractable, or broadly applicable.

@@

1.  Recent years have seen a surge in new approaches for doing so, including ***,*** \_, and \_\_\_\_.
2.  Studies assessing potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
3.  Lagged effects could be different in different habitat types.
4.  This could be a big reason for differences between habitats.
5.  Could also be important for stochastic vs. deterministic models.
6.  Here we...
7.  We conducted these analyses with both deterministic and stochastic IPMs.

@@
